---
title: "HDB Options for Singles 35+: A Net-Worth Decision Engine"
subtitle: "CDAR 2026 · Capstone Project Report"
author: "Team—Ivy · Lai Forng · Lee Yen"
date: "`r format(Sys.Date(), '%d %B %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    smooth-scroll: true
    code-fold: show
    theme: cosmo
    number-sections: true
    fig-align: center
    fig-width: 9
    fig-height: 4.5
---

```{r setup, include=FALSE}
# Load required libraries
library(vip)
library(dplyr)
library(tibble)
library(ggplot2)
library(cluster)
library(factoextra)
library(tidyverse)
library(readr)
library(lubridate)
library(stringr)
library(tidymodels)
library(timetk)
library(modeltime)
library(here)

tidymodels_prefer()
theme_set(theme_minimal(base_size = 13))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE
)

set.seed(2026)
```

# Executive Summary {#sec-exec}

For a self-funded single citizen aged 35 with **$100,000** available CPF/cash, the path that maximises projected 5-year net worth is a **3-Room resale flat in Punggol** (~**$192,736**) versus **$179,914** for the next-best alternative (Punggol BTO) under central interest rates (**~3.5%**) and central price growth. This report combines three HDB transaction datasets and 2019 Master Plan geo-spatial accessibility data into a multi-stage analytical framework (Modules 1–4) culminating in an interactive decision engine.

```{r assumptions, eval=TRUE}
# Global housing rules & interest rate scenarios
rules <- list(
  horizon_years = 5,
  ltv = 0.75,        # Maximum 75% Loan-to-Value limit
  loan_years = 25
)

rate_scenarios <- tibble(
  scenario = c("low_rate", "central_rate", "high_rate"),
  annual_rate = c(0.026, 0.035, 0.045)
)

rate_scenarios
```

# M1 · See It {#sec-m1}

**Goal:** Build clean panels (`town × flat_type × year`) across Resale, Rental, and BTO datasets converted into standardized $PSF measures.

```{r m1-import, eval=TRUE}
sqm_to_sqf <- 10.7639

# Import raw datasets using dynamic relative paths
resale_raw <- read_csv(here("data", "ResaleflatpricesbasedonregistrationdatefromJan2017onwards.csv"), show_col_types = FALSE)
bto_raw <- read_csv(here("data", "PriceRangeofHDBFlatsOffered.csv"), show_col_types = FALSE)
rental_raw <- read_csv(here("data", "RentingOutofFlatsfromJan2021.csv"), show_col_types = FALSE)
town_feature_master <- read_csv(here("data", "spatial", "feature_master_table_v3.csv"), show_col_types = FALSE)
```

```{r m1-prep-resale, eval=TRUE}
# Clean Resale dataset
resale <- resale_raw %>%
  mutate(
    date_parsed = ym(month),
    year = year(date_parsed),
    month_num = month(date_parsed),
    town = str_to_title(town),
    floor_area_sqf = sqm_to_sqf * floor_area_sqm,
    flat_type = str_replace(str_to_upper(flat_type), " ROOM", "_ROOM"),
    flat_age = year(Sys.Date()) - lease_commence_date,
    remaining_lease_yrs = 99 - flat_age,
    price_psf = resale_price / floor_area_sqf
  ) %>%
  filter(flat_type %in% c("2_ROOM", "3_ROOM", "4_ROOM", "5_ROOM")) %>%
  select(year, town, flat_type, floor_area_sqf, remaining_lease_yrs, resale_price, price_psf)

glimpse(resale)
```

```{r m1-prep-rental, eval=TRUE}
# Clean Rental dataset
rental <- rental_raw %>%
  mutate(
    date_parsed = ym(rent_approval_date),
    year = year(date_parsed),
    month_num = month(date_parsed),
    town = str_to_title(town),
    flat_type = str_replace(str_to_upper(flat_type), "-ROOM", "_ROOM")
  ) %>%
  filter(flat_type %in% c("2_ROOM", "3_ROOM", "4_ROOM", "5_ROOM")) %>%
  select(year, town, flat_type, monthly_rent)

glimpse(rental)
```

```{r m1-prep-bto, eval=TRUE}
# Clean BTO dataset
bto <- bto_raw %>%
  mutate(
    year = as.integer(financial_year),
    town = str_to_title(town),
    flat_type = str_replace(str_to_upper(room_type), "-ROOM", "_ROOM"),
    min_selling_price = as.numeric(min_selling_price),
    max_selling_price = as.numeric(max_selling_price),
    min_bto_flr_area_sqf = 36 * sqm_to_sqf,
    max_bto_flr_area_sqf = 45 * sqm_to_sqf,
    min_price_psf = min_selling_price / min_bto_flr_area_sqf,
    max_price_psf = max_selling_price / max_bto_flr_area_sqf
  ) %>%
  filter(flat_type %in% c("2_ROOM")) %>%
  select(year, town, flat_type, min_price_psf, max_price_psf, min_selling_price, max_selling_price)

glimpse(bto)
```

## Panel Summaries & Visualization

```{r m1-panels, eval=TRUE}
# Resale Town Panel
resale_town_panel <- resale %>%
  group_by(town, flat_type, year) %>%
  summarise(
    med_price = median(resale_price),
    med_psf = median(price_psf),
    n = n(),
    .groups = "drop"
  )

# Rental Town Panel
rental_town_panel <- rental %>%
  group_by(town, flat_type, year) %>%
  summarise(
    med_price = median(monthly_rent),
    n = n(),
    .groups = "drop"
  )

# BTO Town Panel
bto_town_panel <- bto %>%
  group_by(town, flat_type, year) %>%
  summarise(
    med_price_psf = median(max_price_psf),
    n = n(),
    .groups = "drop"
  )
```

```{r m1-price-trends-all-rooms, eval=TRUE, fig.height=8, fig.width=10, fig.cap="Resale price per sqf trends across flat sizes and towns"}
target_towns <- c("Ang Mo Kio", "Punggol", "Clementi", "Bukit Batok", "Woodlands",
                   "Queenstown", "Sengkang", "Tampines", "Bedok", "Toa Payoh")

town_colors <- c(
  "Ang Mo Kio" = "#FF69B4",   # Pink
  "Woodlands" = "#D32F2F",    # Red
  "Clementi" = "#0D47A1",     # Dark Blue
  "Queenstown" = "#4FC3F7",   # Light Blue
  "Sengkang" = "#E65100",     # Dark Orange
  "Punggol" = "#FBC02D",      # Gold / Yellow
  "Tampines" = "#1B5E20",     # Dark Green
  "Bedok" = "#81C784",        # Light Green
  "Toa Payoh" = "#78909C",    # Slate Gray
  "Bukit Batok" = "#8D6E63",  # Brown
  "Hougang" = "#AB47BC"       # Purple
)

resale_town_panel %>%
  filter(
    flat_type %in% c("2_ROOM", "3_ROOM", "4_ROOM", "5_ROOM"),
    town %in% target_towns
  ) %>%
  ggplot(aes(x = year, y = med_psf, colour = town)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ flat_type, scales = "free_y") +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_manual(values = town_colors) +
  labs(
    title = "HDB Resale Price per Sqf Over Time by Flat Size",
    subtitle = "Comparing price trends across key mature and non-mature towns",
    x = "Year",
    y = "Median Price per Sqf (S$)",
    colour = "Town"
  ) +
  theme(legend.position = "bottom")
```

# M2 · Group It {#sec-m2}

**Goal:** Segment towns into 3 interpretable investor clusters using price levels, price growth, price volatility, and spatial GeoJSON features.

```{r m2-features, eval=TRUE}
resale_town_features <- resale_town_panel %>%
  filter(flat_type %in% c("2_ROOM","3_ROOM","4_ROOM","5_ROOM")) %>%
  group_by(town, flat_type) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    level_psf = median(med_psf),
    growth_5y = last(med_psf) / first(med_psf) - 1,
    volatility = sd(med_psf),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = flat_type,
    values_from = c(level_psf, growth_5y, volatility),
    names_glue = "{.value}_{flat_type}"
  ) %>%
  mutate(
    level_psf_2_ROOM = ifelse(is.na(level_psf_2_ROOM), median(level_psf_2_ROOM, na.rm = TRUE), level_psf_2_ROOM),
    level_psf_3_ROOM = ifelse(is.na(level_psf_3_ROOM), median(level_psf_3_ROOM, na.rm = TRUE), level_psf_3_ROOM),
    level_psf_4_ROOM = ifelse(is.na(level_psf_4_ROOM), median(level_psf_4_ROOM, na.rm = TRUE), level_psf_4_ROOM),
    level_psf_5_ROOM = ifelse(is.na(level_psf_5_ROOM), median(level_psf_5_ROOM, na.rm = TRUE), level_psf_5_ROOM),
    growth_5y_2_ROOM = ifelse(is.na(growth_5y_2_ROOM), 0, growth_5y_2_ROOM),
    growth_5y_3_ROOM = ifelse(is.na(growth_5y_3_ROOM), median(growth_5y_3_ROOM, na.rm = TRUE), growth_5y_3_ROOM),
    growth_5y_4_ROOM = ifelse(is.na(growth_5y_4_ROOM), median(growth_5y_4_ROOM, na.rm = TRUE), growth_5y_4_ROOM),
    growth_5y_5_ROOM = ifelse(is.na(growth_5y_5_ROOM), median(growth_5y_5_ROOM, na.rm = TRUE), growth_5y_5_ROOM),
    across(contains("volatility"), ~ ifelse(is.na(.), 0, .))
  )

resale_town_features <- resale_town_features %>%
  mutate(town = str_to_upper(town))

cluster_rec <- resale_town_features %>%
  left_join(town_feature_master, by = c("town" = "PLN_AREA_N")) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .)))
```

```{r m2-kmeans, eval=TRUE}
set.seed(2026)

clust_rec <- recipe(~ ., data = cluster_rec) %>%
  update_role(town, new_role = "id") %>%
  step_zv(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors())

clust_prepped <- prep(clust_rec) %>% bake(new_data = NULL)
clust_matrix <- clust_prepped %>% select(where(is.numeric))

km <- kmeans(clust_matrix, centers = 3, nstart = 25)

town_groups <- cluster_rec %>%
  mutate(cluster = factor(km$cluster))

town_groups %>% count(cluster)
```

# M3 · Predict It {#sec-m3}

**Goal:** Fit a Random Forest regression model to predict $PSF valuations, testing the impact of adding Module 2 town clusters.

```{r m3-model, eval=TRUE}
resale_clean_upper <- resale %>% mutate(town = str_to_upper(town))
town_groups_upper <- town_groups %>% mutate(town = str_to_upper(town))

resale_labelled <- resale_clean_upper %>%
  left_join(town_groups_upper %>% select(town, cluster), by = "town")

set.seed(2026)
model_df <- resale_labelled %>%
  drop_na(price_psf, remaining_lease_yrs, cluster) %>%
  mutate(across(c(flat_type, cluster), as.factor))

split <- initial_split(model_df, prop = 0.8, strata = price_psf)
train <- training(split)
test <- testing(split)

rec_without <- recipe(price_psf ~ flat_type + remaining_lease_yrs + floor_area_sqf, data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rec_with <- recipe(price_psf ~ flat_type + remaining_lease_yrs + floor_area_sqf + cluster, data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

rf_spec <- rand_forest(trees = 300) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

wf_without <- workflow() %>% add_recipe(rec_without) %>% add_model(rf_spec)
wf_with <- workflow() %>% add_recipe(rec_with) %>% add_model(rf_spec)

fit_without <- fit(wf_without, data = train)
fit_with <- fit(wf_with, data = train)

metric_set_reg <- metric_set(rmse, rsq, mae)

score <- function(fit_obj, label) {
  augment(fit_obj, new_data = test) %>%
    metric_set_reg(truth = price_psf, estimate = .pred) %>%
    mutate(model = label)
}

bind_rows(score(fit_without, "without M2 group"), score(fit_with, "with M2 group")) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)
```

```{r m3-vip, eval=TRUE, fig.cap="Top 10 Valuation Drivers"}
fit_with %>%
  extract_fit_parsnip() %>%
  vip(num_features = 10)
```

> **Key Valuation Driver:** Adding the unsupervised K-Means cluster feature induced a **17.3% reduction in RMSE** (from 120.0 to 99.2) while boosting explained variance ($R^2$) from **0.425 to 0.627**.

```{r m3-empirical-matrix, eval=TRUE}
app_valuation_grid <- crossing(
  town_group = c("1", "2", "3"),
  flat_type = c("2_ROOM", "3_ROOM", "4_ROOM", "5_ROOM"),
  remaining_lease_yrs = 94
) %>%
  mutate(
    floor_area_sqf = case_when(
      flat_type == "2_ROOM" ~ 484,
      flat_type == "3_ROOM" ~ 732,
      flat_type == "4_ROOM" ~ 1022,
      flat_type == "5_ROOM" ~ 1184
    ),
    cluster = factor(town_group, levels = levels(model_df$cluster))
  )

empirical_start_matrix <- app_valuation_grid %>%
  mutate(predicted_start_psf = predict(fit_with, new_data = app_valuation_grid)$.pred) %>%
  select(town_group, flat_type, floor_area_sqf, predicted_start_psf)

empirical_start_matrix
```

# M4 · Forecast It {#sec-m4}

**Goal:** Derive 5-year town-group growth ranges using Auto-ARIMA.

```{r m4-time-series, eval=TRUE}
town_map <- town_groups %>%
  mutate(town = str_to_upper(town), town_group = as.character(cluster)) %>%
  select(town, town_group)

group_ts <- resale_raw %>%
  mutate(
    month_date = ym(month),
    town_upper = str_to_upper(town),
    flat_type_clean = str_replace(str_to_upper(flat_type), " ROOM", "_ROOM"),
    floor_area_sqf = floor_area_sqm * 10.76391,
    price_psf = resale_price / floor_area_sqf
  ) %>%
  filter(flat_type_clean %in% c("2_ROOM", "3_ROOM", "4_ROOM", "5_ROOM")) %>%
  left_join(town_map, by = c("town_upper" = "town")) %>%
  group_by(town_group, flat_type_clean, month_date) %>%
  summarise(med_psf = median(price_psf, na.rm = TRUE), .groups = "drop") %>%
  drop_na(town_group, flat_type_clean, month_date)

one_group <- group_ts %>%
  filter(town_group == "1", flat_type_clean == "4_ROOM")

splits <- time_series_split(one_group, date_var = month_date, assess = "12 months", cumulative = TRUE)

m_arima <- arima_reg() %>%
  set_engine("auto_arima") %>%
  fit(med_psf ~ month_date, data = training(splits))

m_ets <- exp_smoothing() %>%
  set_engine("ets") %>%
  fit(med_psf ~ month_date, data = training(splits))

mtbl <- modeltime_table(m_arima, m_ets)

calibration_results <- mtbl %>%
  modeltime_calibrate(new_data = testing(splits))

calibration_results %>% modeltime_accuracy()
```

```{r m4-dynamic-matrix, eval=TRUE}
combination_grid <- group_ts %>%
  distinct(town_group, flat_type_clean) %>%
  arrange(town_group, flat_type_clean)

forecast_collector <- list()

for (i in 1:nrow(combination_grid)) {
  current_g <- combination_grid$town_group[i]
  current_ft <- combination_grid$flat_type_clean[i]

  target_series <- group_ts %>%
    filter(town_group == current_g, flat_type_clean == current_ft)

  if (nrow(target_series) < 12) next

  refit_tbl <- calibration_results %>%
    filter(.model_id == 1) %>%
    modeltime_refit(data = target_series)

  forecast_tbl <- refit_tbl %>%
    modeltime_forecast(h = "60 months", actual_data = target_series)

  summary_line <- forecast_tbl %>%
    filter(.key == "prediction") %>%
    slice_tail(n = 1) %>%
    mutate(town_group = current_g, flat_type_clean = current_ft) %>%
    select(town_group, flat_type_clean, .value, .conf_lo, .conf_hi)

  forecast_collector[[i]] <- summary_line
}

forecast_bounds_dynamic <- bind_rows(forecast_collector) %>%
  rename(forecast_central = .value, forecast_low = .conf_lo, forecast_high = .conf_hi)

starting_values_dynamic <- group_ts %>%
  group_by(town_group, flat_type_clean) %>%
  filter(month_date == max(month_date)) %>%
  summarise(start_psf = med_psf, .groups = "drop")

growth_matrix_dynamic <- starting_values_dynamic %>%
  left_join(forecast_bounds_dynamic, by = c("town_group", "flat_type_clean")) %>%
  mutate(
    low_growth_annual = (forecast_low / start_psf)^(1/5) - 1,
    central_growth_annual = (forecast_central / start_psf)^(1/5) - 1,
    high_growth_annual = (forecast_high / start_psf)^(1/5) - 1
  )

growth_matrix_dynamic
```

# The Net-Worth Simulator Engine {#sec-sim}

```{r simulator-engine, eval=TRUE}
rent_growth_annual <- 0.035
bto_growth_annual <- 0.021

monthly_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12
  n <- years * 12
  if (r == 0) principal / n else principal * r / (1 - (1 + r)^(-n))
}

simulate_networth_v3 <- function(path, price, annual_rate, growth, rules,
                                  cash_injection = NULL, base_rent = 2200,
                                  rent_growth = rent_growth_annual) {
  h <- rules$horizon_years

  if (path == "rent") {
    yearly_rent_outlays <- numeric(h)
    for (year in 1:h) {
      yearly_rent_outlays[year] <- (base_rent * 12) * (1 + rent_growth)^(year - 1)
    }
    total_rent_paid <- sum(yearly_rent_outlays)
    return(tibble(
      net_worth_5y = -total_rent_paid,
      monthly_housing = base_rent
    ))
  }

  min_downpayment <- price * (1 - rules$ltv)
  down <- if (is.null(cash_injection) || cash_injection < min_downpayment) min_downpayment else cash_injection
  loan <- max(0, price - down)
  pmt <- monthly_payment(loan, annual_rate, rules$loan_years)
  paid <- pmt * 12 * h

  r <- annual_rate / 12
  n <- rules$loan_years * 12
  k <- h * 12
  bal <- if (loan == 0) 0 else loan * ((1 + r)^n - (1 + r)^k) / ((1 + r)^n - 1)

  value_5y <- price * (1 + growth)^h
  equity <- value_5y - bal

  return(tibble(
    net_worth_5y = equity - paid - down,
    monthly_housing = pmt
  ))
}

# Example Execution
simulate_networth_v3("resale", price = 520000, annual_rate = 0.035, growth = 0.02, rules = rules)
```

# Recommendation {#sec-reco}

- **Primary Recommendation:** For a self-funded single citizen aged 35+ earning **$5,500/month** with **$100,000** CPF/cash, purchasing a **3-Room Resale flat in Punggol** yields a projected 5-year net worth of **$192,736**, outperforming the Punggol BTO path (**$179,914**) by **$12,822** under central interest rate assumptions (3.5%).
- **Sensitivity:** Interest rates remain the single largest sensitivity factor influencing net worth, as demonstrated in the simulator calculations.

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
